---
title: "Reproducible Pitch Presentation"
author: "Alberto Rossi"
date: "17/06/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Heartbeat Estimator

This is a simple application calculates the number of times a heart beats, considering an average of 60 beats per minute.


[Heart Estimator](https://albrossi.shinyapps.io/DataProductApp/)


## Application structure

- The application is done in Shiny.
- Uses a slicer with the measurement in Years, available in the sidebarPanel.
- The result is presented in the mainPanel, with the estimated heart rate calculation * age informed.

## Calculation

The calculation is made by multiplying the age by the heart rate estimate - which is, on average, used here, 60 per minute.

A format command was used to better display the decimal places, since they are very large numbers!

```{r echo = TRUE}
    #totalbeats <- input$beats * 31536000
    #format(totalbeats, format="d", big.mark=',')
```

# Thank you
