---
title: "Machine Learning - Final Test"
author: "Alberto Rossi"
date: "16/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 

They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


# Objective

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. 

You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


# Data loading and preparing
  
Reading of data according to informed URLs and loading of files in data frames.

Once this is done, some data (NearZero, NAs) and some columns that bring irrelevant information to the analysis are cleaned up.

This step is critical to the downsizing of the data and better performance execution analysis.

```{r}
  library(caret)
  
  #Reading and confirming data
  fnameTraining <- "pml-training.csv"
  fnameTesting <- "pml-testing.csv"
  
  fileURLTraining <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  fileURLTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  
  # File already exists?
  if (!file.exists(fnameTraining)){
    download.file(url = fileURLTraining, 
                  destfile = fnameTraining)
  }  
  
  # File already exists?
  if (!file.exists(fnameTesting)){
    download.file(url = fileURLTest, 
                  destfile = fnameTesting)
  }  
  
  # Data read
  trainingData <- read.table(fnameTraining,
                             sep = ",",
                             dec = ".",
                             header = TRUE,
                             na.strings = "NA") 

  validationData <- read.table(fnameTesting,
                             sep = ",",
                             dec = ".",
                             header = TRUE,
                             na.strings = "NA")
  
  # Checking datasets
  
  str(trainingData)
  str(validationData)
  
  # Cleaning: Near Zero Variance variables from datasets
  nzv <- nearZeroVar(trainingData, saveMetrics = T)
  trainingData <- trainingData[, !nzv$nzv] 
  validationData <- validationData[, !nzv$nzv]
  
  # Cleaning: Remove columns that contain NA's
  nas <- (colSums(is.na(trainingData)) == 0)
  trainingData <- trainingData[, nas]
  validationData <- validationData[, nas]
  
  # Cleaning: remove 1 to 5 columns from datasets (useless)
  trainingData <- trainingData[ , -c(1:5)]
  validationData <- validationData[ , -c(1:5)]

```

# Data partition

We will partition the data according to the rule below:

* Training data = 60% training file
* Testing data = 40% training file
* Validation data = 100% original testing file

```{r}
  set.seed(1609)
  partitionIndex <- createDataPartition(y=trainingData$classe, p=0.6, list=FALSE)
  trainingData <- trainingData[partitionIndex, ]
  testingData <- trainingData[-partitionIndex, ]

```

# Prediction models selection 

Lets test 3 different models to see the best out-of-sample accuracy:

* Random forest decision trees (rf)
* Stochastic gradient boosting trees (gbm)
* Linear Discriminant Analysis (lda)

In this exercise we will use explicit parallel processing via R, to improve the training time of the data.

```{r}
  # Training our data
  
  
  load(file = "C:/RStudio/datasciencecoursera/Modelrf.RData")
  #model_rf <- train(classe ~ ., 
  #                   data=trainingData,
  #                  method='rf')
  #save(model_rf, file="./Modelrf.RData")
  
  load(file = "C:/RStudio/datasciencecoursera/Modelgbm.RData")
  #model_gbm <- train(classe ~ .,
  #                   data=trainingData,
  #                   method='gbm')
  #save(model_gbm, file="./Modelgbm.RData")
  
  load(file = "C:/RStudio/datasciencecoursera/Modelda.RData")
  #model_lda <- train(classe ~ .,
  #                  data=trainingData,
  #                  method='lda')
  #save(model_lda, file="./Modelda.RData")
```

# Model analysis

After running the 3 models, we will perform the cross validation and compare the accuracy of the models, choosing the best one to perform the validation step.

```{r}
  # Run RF model
  predmodel_rf <- predict(model_rf, newdata=testingData)
  cmmodel_rf <- confusionMatrix(table(predmodel_rf, testingData$classe))
  
  # Run GBM model
  predmodel_gbm <- predict(model_gbm, newdata=testingData)
  cmmodel_gbm <- confusionMatrix(table(predmodel_gbm, testingData$classe))
 
  # Run LDA model 
  predmodel_lda <- predict(model_lda, newdata=testingData)
  cmmodel_lda <- confusionMatrix(table(predmodel_lda, testingData$classe))
  
  accuracyTable <- data.frame(Model = c('RF', 'GBM', 'LDA'),
                              Accuracy = rbind(cmmodel_rf$overall[1],
                                               cmmodel_gbm$overall[1],
                                               cmmodel_lda$overall[1]))
  
  print(accuracyTable)
 

```

# Model Prediction
  
The best model is choosen to predict values from Validation dataset, with 20 observations.
In this case, RF:

```{r}
  predData <- predict(model_rf, newdata=validationData)
  PredResults <- data.frame(problem_id=validationData$problem_id,
                                            predicted=predData)
  print(PredResults)
```


# Conclusion

The Random Forest model delivers the best analysis accuracy for the data in question.

```{r}
  varImpObj <- varImp(model_rf)
  plot(varImpObj, main = "Important Variables - RF", top = 20)
```
