---
title: "Regression Models - Final Project"
author: "Alberto Rossi"
date: "13/06/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

This is a statistical study on exploring the relationship between a set of variables and miles per gallon (MPG) (outcome).

The goal is to get the following answers:

## 1. Is an automatic or manual transmission better for MPG?

## 2. Quantify the MPG difference between automatic and manual transmissions


# Exploratory data analysis

The data provided are those of Mtcars. No download is needed.

```{r}
  str(mtcars)
  mtcars
```

The questions refer to basically two variables:

* mpg: Miles per gallon
* am: Transmission (0 = automatic, 1 = manual)

```{r}
  mpg<- mtcars$mpg
  am <- mtcars$am
  unique(mpg)
  unique(am)
```

## Results:
No missing values, no gaps - good.

Now, lets go for some answers.


# 1. Is an automatic or manual transmission better for MPG?
   
The understanding of this question will be assumed as follows:
- Which type of transmission presented indicates the lowest mpg index;
- That is, which transmission has the best cost-benefit ratio in terms of fuel consumption (mpg).
  
To better visualize some pattern on mpg and transmission, we will plot a graph for better reading:

```{r}
  boxplot(mpg~am,
          ylab = "MPG",
          xlab = "Gear | 0-Auto  1-Manual",
          main="MPG vs Gears")

```

Let's try to find some correlation via Linear Regression. Since the variables are numeric, we can use them directly, without "Factoring".

```{r}
 linearR <- lm(mpg~am)
  plot(mpg~am,
       ylab = "MPG",
       xlab = "Gear | 0-Auto  1-Manual",
       main="MPG vs Gears")
  abline(linearR, lwd = 1)
```
 
Check coeficients summary: 

```{r}
  summary(linearR)
```

## Results:

Coefficient of 7.245, with p-value of 0.000285 (***) - quite significant.
There is some relationship between MPG and transmission. But how much?
  
  
# 2. Quantify the MPG difference between automatic and manual transmissions
  
Let's now analyze the influence of the transmission on the MPG:

```{r}
  summary(linearR)
```

Multiple R-squared is 0.3598, meaning that just a part of MPG can be attributed to transmission. 

To discover more information about the Variance:

```{r}
  linearRV <- aov(mpg ~ ., data = mtcars)
  summary(linearRV)
```  

Again, lets focus on itens with significance: 
* *** cyl = Number of cylinders
* * disp = Displacement
* wt = Weight (1000 lbs)

```{r}
  linearRVF <- lm(mpg ~ cyl + disp + wt + am, data = mtcars)
  summary(linearRVF)
```  
 
## Results:
This analysis show an Multiple R-squared:  0.8327, suggesting that +83% of variance can be explained by cyl, disp, wt and transmission. 

P-values for cyl and weight are below 0.5, suggesting that these are confounding variables in the relation between car Transmission Type and Miles per Gallon.
  
Now lets check in more detail residuals:

```{r}
  par(mfrow = c(2, 2))
  plot(linearRVF)
```  

## Results:

The “Residuals vs Fitted” plot here shows us that the residuals are homoscedastic. 

We can also see that they are normally distributed, with the exception of a few outliers.
  

# Conclusions

Based on this analysis we can conclude:

* Manual transmission get more miles per gallon compared against Automatic transmission.
* Mpg decreases negligibly with increase of hp.
* If number of cylinders increases, mpg will decrease.

